---
title: "Final Project"
author: "Tourigny, Timmy; Misiorek, Ty; Chandra, Arnav"
date: 'Dec 6, 2024'
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
editor_options: 
  markdown: 
    wrap: 72
---

**APMA 3150 \| Fall 2024 \| University of Virginia**

***"On my honor, I pledge that I have neither given nor received unauthorized aid on this assignment." - The author of this RMarkdown file.***

<!--- Solution Region --->

```{css solution-region, echo=FALSE}
.solution {
    background-color: #232D4B10;
    border-style: solid;
    border-color: #232D4B;
    padding: .5em;
    margin: 20px
}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



# [Analysis]{style="color:#FF7F50"}

Data Retrieval

```{r}
library(ggplot2)
library(dplyr)

#remember to set working directory
data <- read.csv("death-rates-from-air-pollution.csv")
# Rename columns for simplicity
colnames(data) <- c("Entity", "Code", "Year", "Total_Deaths", "Indoor_Deaths", "Outdoor_PM_Deaths", "Outdoor_Ozone_Deaths")

data <- data[!grepl("SDI", data$Entity), ]
data <- data[!grepl("income", data$Entity), ]


```


```{r}
aggregated_data <- aggregate(
  cbind(Total_Deaths, Indoor_Deaths, Outdoor_PM_Deaths, Outdoor_Ozone_Deaths) ~ Year,
  data = data,
  sum,
  na.rm = TRUE
)

model1 <- lm(aggregated_data$Total_Deaths ~ aggregated_data$Year)
plot(aggregated_data$Year, aggregated_data$Total_Deaths,
     xlab = "Year",
     ylab = "Total Deaths (per 100,000 people)",
     main = "Global Trend of Total Deaths Due to Air Pollution (1990-2017)"
)
abline(model1, col = "blue")
model1$coefficients[2]
summary(model1)$r.squared

plot(aggregated_data$Year, aggregated_data$Indoor_Deaths,
     type = "l",
     col = "red",
     lwd = 2,
     xlab = "Year",
     ylab = "Deaths (per 100,000 people)",
     main = "Global Trends of Air Pollution Deaths by Type (1990-2017)",
     ylim = c(0,16000)
)
lines(aggregated_data$Year, aggregated_data$Outdoor_PM_Deaths, col = "green", lwd = 2)
lines(aggregated_data$Year, aggregated_data$Outdoor_Ozone_Deaths, col = "blue", lwd = 2)

legend("topright",
       legend = c("Indoor Pollution", "Outdoor PM Pollution", "Outdoor Ozone Pollution"),
       col = c("red", "green", "blue"),
       lwd = 2,
       bty = "n")

cor(aggregated_data$Year,aggregated_data$Indoor_Deaths)
cor(aggregated_data$Year,aggregated_data$Outdoor_PM_Deaths)
cor(aggregated_data$Year,aggregated_data$Outdoor_Ozone_Deaths)


```

```{r}
aggregated_data <- aggregate(
  cbind(Total_Deaths, Indoor_Deaths, Outdoor_PM_Deaths, Outdoor_Ozone_Deaths) ~ Year,
  data = data,
  sum,
  na.rm = TRUE
)

#fit lm for total deaths and years
model1 <- lm(aggregated_data$Total_Deaths ~ aggregated_data$Year)
plot(aggregated_data$Year, aggregated_data$Total_Deaths,
     xlab = "Year",
     ylab = "Total Deaths (per 100,000 people)",
     main = "Global Trend of Total Deaths Due to Air Pollution (1990-2017)",
     col = "blue"
)
abline(model1,)
cat("Total Deaths Slope:", model1$coefficients[2], "\n")
cat("R-squared for Total Deaths:", summary(model1)$r.squared, "\n")

#fit for each subset of deaths
model_indoor <- lm(aggregated_data$Indoor_Deaths ~ aggregated_data$Year)
model_pm <- lm(aggregated_data$Outdoor_PM_Deaths ~ aggregated_data$Year)
model_ozone <- lm(aggregated_data$Outdoor_Ozone_Deaths ~ aggregated_data$Year)

plot(aggregated_data$Year, aggregated_data$Indoor_Deaths,
     lwd = 2,
     xlab = "Year",
     ylab = "Deaths (per 100,000 people)",
     main = "Global Trends of Air Pollution Deaths by Type (1990-2017)",
     ylim = c(0, 16000),
     col = "red"
)
points(aggregated_data$Year, aggregated_data$Outdoor_PM_Deaths, lwd = 2, col = "green")
points(aggregated_data$Year, aggregated_data$Outdoor_Ozone_Deaths, lwd = 2, col = "blue")

#add regression lines
abline(model_indoor, )
abline(model_pm,)
abline(model_ozone, )

legend("topright",
       legend = c("Indoor Pollution", "Outdoor PM Pollution", "Outdoor Ozone Pollution"),
       col = c("red", "green", "blue"),
       lwd = 2,
       lty = c(1, 1, 1),
       bty = "n")

#display regression slopes and R-squared values
cat("Indoor Pollution Slope:", model_indoor$coefficients[2], "\n")
cat("R-squared for Indoor Pollution:", summary(model_indoor)$r.squared, "\n")
cat("Outdoor PM Pollution Slope:", model_pm$coefficients[2], "\n")
cat("R-squared for Outdoor PM Pollution:", summary(model_pm)$r.squared, "\n")
cat("Outdoor Ozone Pollution Slope:", model_ozone$coefficients[2], "\n")
cat("R-squared for Outdoor Ozone Pollution:", summary(model_ozone)$r.squared, "\n")


```




```{r}
#plotting number of deaths (per 100k people) by year for a select few countries to see how trends changed throughout different regions and if we see that
#most regions are pretty similar to each other or if they are kind of independent

countries <- c("United States", "Afghanistan", "China", "United Arab Emirates")

for (country in countries) {
  country_data <- data[data$Entity == country, ]
  
  plot(
    country_data$Year, country_data$Total_Deaths, 
    type = "o",
    col = "blue",
    pch = 16,
    lwd = 2,
    xlab = "Year", 
    ylab = "Total Deaths (per 100,000)",
    main = paste("Total Deaths Due to Pollution in", country),
    cex.main = 1.5,
    cex.lab = 1.2,  
    cex.axis = 1 
  )
}
```

```{r}
#From the first couple of countries we looked at, we observed a downwards trend in deaths, so now lets see if we can identify any upwards trending countries. Also determine what the average change in deaths is in the dataset so we can compare these against the average

unique_countries <- unique(data$Entity)
slopes <- numeric(length(unique_countries))
sum_slopes <- 0
count_slopes <- 0

for (i in seq_along(unique_countries)) {
  country_data <- data[data$Entity == unique_countries[i], ]
  #fit lm if enough dawta for that country
  if (nrow(country_data) > 1) {
    model <- lm(Total_Deaths ~ Year, data = country_data)
    slopes[i] <- coef(model)["Year"]
    sum_slopes <- sum_slopes + slopes[i]
    count_slopes <- count_slopes + 1
  } else {
    slopes[i] <- NA
  }
}

#identify upward trending countries
results <- data.frame(Country = unique_countries, Slope = slopes)
upward_trend <- results[results$Slope > 0, ]
upward_trend <- upward_trend[order(-upward_trend$Slope), ]

#create averages
average_slope <- sum_slopes / count_slopes

upward_trend
cat("Average Slope:", average_slope)
```



```{r}
#plot the upwards trend countries ()
upward_countries <- c("Zimbabwe", "Lesotho", "Philippines", "Libya", "Uzbekistan")
upward_countries_plot <- head(upward_countries,2)
slope_sorted_countries <- results[order(results$Slope),]
downward_countries_plot <- slope_sorted_countries[1:2, ]
average_countries_plot <- slope_sorted_countries[length(slope_sorted_countries[,1])/2:length(slope_sorted_countries[,1])/2+1,]

plot_country_trend <- function(country) {
  country_data <- data[data$Entity == country, ]
  
  plot(
    country_data$Year, country_data$Total_Deaths, 
    type = "o",
    col = "blue",
    pch = 16,
    lwd = 2,
    xlab = "Year", 
    ylab = "Total Deaths (per 100,000)",
    main = paste(country),
    cex.main = 1.5,
    cex.lab = 1.2,  
    cex.axis = 1 
  )
}

par(mfrow = c(3,2))
plot_country_trend(downward_countries_plot[1,1])
plot_country_trend(downward_countries_plot[2,1])
plot_country_trend(average_countries_plot[1,1])
plot_country_trend(average_countries_plot[2,1])
plot_country_trend(upward_countries_plot[1])
plot_country_trend(upward_countries_plot[2])


```


Most countries have similar downward trend, except for the high_slope outliers, which have a spike around 2010

now, perform k-means clustering to group data

```{r}
developing_countries <- c(
  "Afghanistan", "Angola", "Bangladesh", "Benin", "Bhutan", 
  "Bolivia", "Botswana", "Burkina Faso", "Burundi", "Cambodia", 
  "Cameroon", "Central African Republic", "Chad", "Comoros", 
  "Congo", "Cote d'Ivoire", "Democratic Republic of Congo", "Djibouti", 
  "Eritrea", "Ethiopia", "Gambia", "Ghana", "Guatemala", "Guinea", 
  "Guinea-Bissau", "Haiti", "Honduras", "India", "Indonesia", 
  "Kenya", "Kiribati", "Kyrgyzstan", "Laos", "Lesotho", "Liberia", 
  "Madagascar", "Malawi", "Mali", "Mauritania", "Micronesia (country)", 
  "Moldova", "Mozambique", "Myanmar", "Namibia", "Nepal", 
  "Nicaragua", "Niger", "Nigeria", "Pakistan", "Papua New Guinea", 
  "Philippines", "Rwanda", "Samoa", "Sao Tome and Principe", "Senegal", 
  "Sierra Leone", "Solomon Islands", "Somalia", "South Sudan", 
  "Sudan", "Swaziland", "Tajikistan", "Tanzania", "Timor", 
  "Togo", "Uganda", "Uzbekistan", "Vanuatu", "Vietnam", "Yemen", 
  "Zambia", "Zimbabwe"
)

data <- data %>%
  mutate(Status = ifelse(Entity %in% developing_countries, "Developing", "Developed"))

t_test_result <- t.test(
  Total_Deaths ~ Status, 
  data = data, 
  var.equal = FALSE, 
  na.action = na.omit
)
print(t_test_result)

```
The t-test result suggests that is it almost certain that developing countries are more likely to have higher total death by pollution. 



```{r}
#split on median for high vs low death count
data <- data %>%
  mutate(High_Deaths = ifelse(Total_Deaths > median(Total_Deaths), "High", "Low"))

# make contingency table and chi square
contingency_table <- table(data$Status, data$High_Deaths)

# Perform the chi-square test
chi_square_result <- chisq.test(contingency_table)

developed_low <- contingency_table["Developed", "Low"]
developed_high <- contingency_table["Developed", "High"]
developing_low <- contingency_table["Developing", "Low"]
developing_high <- contingency_table["Developing", "High"]

cat("\nChi-Square Test Results:\n")
cat("=====================================\n")
cat("Contingency Table:\n")
cat("                 Low     High\n")
cat("Developed:     ", developed_low, "    ", developed_high, "\n")
cat("Developing:    ", developing_low, "    ", developing_high, "\n")
cat("\nChi-Square Statistic:      ", round(chi_square_result$statistic, 3), "\n")
cat("Degrees of Freedom:        ", chi_square_result$parameter, "\n")
cat("P-value:                   ", format.pval(chi_square_result$p.value, digits = 3), "\n")
cat("=====================================\n")


```


```{r}
data$Status <- ifelse(data$Entity %in% developing_countries, "Developing", "Developed")

developing_data <- data[data$Status == "Developing", ]

indoor_deaths <- sum(developing_data$Indoor_Deaths, na.rm = TRUE)
outdoor_deaths <- sum(developing_data$Outdoor_PM_Deaths + developing_data$Outdoor_Ozone_Deaths, na.rm = TRUE)

pollution_deaths <- data.frame(
  Type = c("Indoor", "Outdoor"),
  Deaths = c(indoor_deaths, outdoor_deaths)
)

pollution_deaths

```