---
title: "Final Project"
author: "Tourigny, Timmy; Misiorek, Ty; Chandra, Amav"
date: 'Dec 6, 2024'
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
editor_options: 
  markdown: 
    wrap: 72
---

**APMA 3150 \| Fall 2024 \| University of Virginia**

***"On my honor, I pledge that I have neither given nor received unauthorized aid on this assignment." - The author of this RMarkdown file.***

<!--- Solution Region --->

```{css solution-region, echo=FALSE}
.solution {
    background-color: #232D4B10;
    border-style: solid;
    border-color: #232D4B;
    padding: .5em;
    margin: 20px
}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



# [Analysis]{style="color:#FF7F50"}

Data Retrieval

```{r}
library(ggplot2)
library(dplyr)

#remember to set working directory
data <- read.csv("death-rates-from-air-pollution.csv")
# Rename columns for simplicity
colnames(data) <- c("Entity", "Code", "Year", "Total_Deaths", "Indoor_Deaths", "Outdoor_PM_Deaths", "Outdoor_Ozone_Deaths")


```


```{r}
aggregated_data <- aggregate(
  cbind(Total_Deaths, Indoor_Deaths, Outdoor_PM_Deaths, Outdoor_Ozone_Deaths) ~ Year,
  data = data,
  sum,
  na.rm = TRUE
)

plot(aggregated_data$Year, aggregated_data$Total_Deaths,
     type = "l",
     col = "blue",
     lwd = 2,
     xlab = "Year",
     ylab = "Total Deaths (per 100,000 people)",
     main = "Global Trend of Total Deaths Due to Air Pollution (1990-2017)"
)

plot(aggregated_data$Year, aggregated_data$Indoor_Deaths,
     type = "l",
     col = "red",
     lwd = 2,
     xlab = "Year",
     ylab = "Deaths (per 100,000 people)",
     main = "Global Trends of Air Pollution Deaths by Type (1990-2017)"
)
lines(aggregated_data$Year, aggregated_data$Outdoor_PM_Deaths, col = "green", lwd = 2)
lines(aggregated_data$Year, aggregated_data$Outdoor_Ozone_Deaths, col = "blue", lwd = 2)

legend("topright",
       legend = c("Indoor Pollution", "Outdoor PM Pollution", "Outdoor Ozone Pollution"),
       col = c("red", "green", "blue"),
       lwd = 2,
       bty = "n")


```




```{r}
#plotting number of deaths (per 100k people) by year for a select few countries to see how trends changed throughout different regions and if we see that
#most regions are pretty similar to each other or if they are kind of independent

countries <- c("United States", "Afghanistan", "China", "United Arab Emirates")

for (country in countries) {
  country_data <- data[data$Entity == country, ]
  
  plot(
    country_data$Year, country_data$Total_Deaths, 
    type = "o",
    col = "blue",
    pch = 16,
    lwd = 2,
    xlab = "Year", 
    ylab = "Total Deaths (per 100,000)",
    main = paste("Total Deaths Due to Pollution in", country),
    cex.main = 1.5,
    cex.lab = 1.2,  
    cex.axis = 1 
  )
}
```

```{r}
#From the first couple of countries we looked at, we observed a downwards trend in deaths, so now lets see if we can identify any upwards trending countries. Also determine what the average change in deaths is in the dataset so we can compare these against the average

unique_countries <- unique(data$Entity)
slopes <- numeric(length(unique_countries))
sum_slopes <- 0
count_slopes <- 0

for (i in seq_along(unique_countries)) {
  country_data <- data[data$Entity == unique_countries[i], ]
  
  # Fit a linear regression if there is sufficient data
  if (nrow(country_data) > 1) {
    model <- lm(Total_Deaths ~ Year, data = country_data)
    slopes[i] <- coef(model)["Year"]
    sum_slopes <- sum_slopes + slopes[i]
    count_slopes <- count_slopes + 1
  } else {
    slopes[i] <- NA
  }
}

#identify upward trending countries
results <- data.frame(Country = unique_countries, Slope = slopes)
upward_trend <- results[results$Slope > 0, ]
upward_trend <- upward_trend[order(-upward_trend$Slope), ]

#create averages
average_slope <- sum_slopes / count_slopes

upward_trend
cat("Average Slope:", average_slope)
```
The average slope of the deaths is -1.612, which is very different from the five we identified with a positive slope. Calculate some z-scores to see how different these are from the mean
```{r}
valid_slopes <- slopes[!is.na(slopes)]
mean_slope <- mean(valid_slopes)
sd_slope <- sd(valid_slopes)

upward_trend$Z_Score <- (upward_trend$Slope - mean_slope) / sd_slope

cat("Average Slope:", mean_slope, "\n")
cat("Standard Deviation of Slopes:", sd_slope, "\n")
upward_trend
```

Check for normality with a qqplot:
```{r}
qqnorm(valid_slopes, main = "QQ Plot of Slopes")
qqline(valid_slopes, col = "red", lwd = 2)

hist(valid_slopes, breaks = 20, probability = TRUE, 
     main = "Histogram of Slopes", 
     xlab = "Slopes", 
     col = "green", 
     border = "black")
lines(density(valid_slopes), col = "darkblue", lwd = 2)
```
The distribution of slopes (change in death rates) is somewhat normally distributed, although we have a large leftward skew


```{r}
#plot the upwards trend countries ()
upward_countries <- c("Zimbabwe", "Lesotho", "Philippines", "Libya", "Uzbekistan")

for (country in upward_countries) {
  country_data <- data[data$Entity == country, ]
  
  plot(
    country_data$Year, country_data$Total_Deaths, 
    type = "o",
    col = "blue",
    pch = 16,
    lwd = 2,
    xlab = "Year", 
    ylab = "Total Deaths (per 100,000)",
    main = paste("Total Deaths Due to Pollution in", country),
    cex.main = 1.5,
    cex.lab = 1.2,  
    cex.axis = 1 
  )
}

#lesotho is landlocked by south africa, so I want to see how this looks in South Africa
country_data <- data[data$Entity == "South Africa", ]
plot(
    country_data$Year, country_data$Total_Deaths, 
    type = "o",
    col = "blue",
    pch = 16,
    lwd = 2,
    xlab = "Year", 
    ylab = "Total Deaths (per 100,000)",
    main = paste("Total Deaths Due to Pollution in South Africa"),
    cex.main = 1.5,
    cex.lab = 1.2,  
    cex.axis = 1 
  )

```


Now that we've identified some countries that have a trend different from expected, compare neighboring countries of these. Classifying these countries manually seemed a lot easier than other methods, so going to do that. However, they were not linear relationships and Libya was really the only one that was still increasing. So instead of continuing with linear regression, gonna do clustering instead
```{r}
neighbors <- list(
  "Libya" = c("Tunisia", "Algeria", "Egypt", "Niger", "Chad", "Sudan"),
  "Zimbabwe" = c("Zambia", "Mozambique", "Namibia", "Botswana", "South Africa"),
  "Lesotho" = c("South Africa", "Eswatini"),
  "Uzbekistan" = c("Turkmenistan", "Afghanistan", "Tajikistan", "Kyrgyzstan", "Kazakhstan"),
  "Philippines" = c("Malaysia", "Brunei", "Taiwan", "Indonesia")
)

all_countries <- unique(c(upward_trend$Country, unlist(neighbors)))
cluster_data <- data[data$Entity %in% all_countries, ]

cluster_features <- cluster_data[, c("Indoor_Deaths", "Outdoor_PM_Deaths", "Outdoor_Ozone_Deaths")]
cluster_features <- na.omit(cluster_features)

cluster_features_scaled <- scale(cluster_features)


set.seed(123)
kmeans_result <- kmeans(cluster_features_scaled, centers = 3)
cluster_data$Cluster <- kmeans_result$cluster

table(cluster_data$Entity, cluster_data$Cluster)

aggregate(cluster_features, by = list(Cluster = kmeans_result$cluster), mean)

```
In this data, we create 3 clusters, cluster 1 representing countries with predominantly Indoor pollution related deaths, cluster 2 representing more Outdoor related, and 3 representing a mixture of both. The goal here was to see if neighboring countries were in the same clusters as the ones who had an upwards trend. 
Zambia, Mozambique", "Namibia", "Botswana", "South Africa

Libya (Cluster 3): Same cluster as 2/6 neighbors (Tunisia, Algeria) Different from 4/6 (Niger, Chad, Egypt, Sudan)
Zimbabwe (Cluster 1): Same cluster as 3/5 (Mozambique, Zambia, Namibia), different from 2/5 (Botswana, South Africa)
Lesotho (Cluster 1): Same cluster as 0/1 (South Africa) (Eswatini not in dataset)
Uzbekistan (Cluster 2): Same cluster as 1/5 (Turkmenistan) different from 4/5 (Afghanistan, Tajikistan, Kyrgyzstan, and Kazakhstan)
Phillippines (Cluster 3): Same cluster as 4/4 (Malaysia, Brunei, Taiwan, Indonesia)

```{r}
# Ensure Cluster is a factor
cluster_data$Cluster <- as.factor(cluster_data$Cluster)

# ANOVA for Indoor Pollution Deaths
anova_indoor <- aov(Indoor_Deaths ~ Cluster, data = cluster_data)
summary(anova_indoor)

# Post-hoc test for Indoor Pollution
if (summary(anova_indoor)[[1]][["Pr(>F)"]][1] < 0.05) {
  print(TukeyHSD(anova_indoor))
}

# ANOVA for Outdoor PM Pollution Deaths
anova_pm <- aov(Outdoor_PM_Deaths ~ Cluster, data = cluster_data)
summary(anova_pm)

# Post-hoc test for Outdoor PM Pollution
if (summary(anova_pm)[[1]][["Pr(>F)"]][1] < 0.05) {
  print(TukeyHSD(anova_pm))
}

# ANOVA for Outdoor Ozone Pollution Deaths
anova_ozone <- aov(Outdoor_Ozone_Deaths ~ Cluster, data = cluster_data)
summary(anova_ozone)

# Post-hoc test for Outdoor Ozone Pollution
if (summary(anova_ozone)[[1]][["Pr(>F)"]][1] < 0.05) {
  print(TukeyHSD(anova_ozone))
}
```
probably going to remove this cell from the analysis, it just states that the different clusters are significantly different from each other, which is pretty obvious and would kind of make our results useless if there was any doubt that this was not the case


```{r}
data$Neighbor_Indoor <- NA
data$Neighbor_PM <- NA
data$Neighbor_Ozone <- NA

for (country in names(neighbors)) {
  neighbor_list <- neighbors[[country]]
  
  neighbor_data <- data[data$Entity %in% neighbor_list, ]
  
  if (nrow(neighbor_data) > 0) {
    data$Neighbor_Indoor[data$Entity == country] <- mean(neighbor_data$Indoor_Deaths, na.rm = TRUE)
    data$Neighbor_PM[data$Entity == country] <- mean(neighbor_data$Outdoor_PM_Deaths, na.rm = TRUE)
    data$Neighbor_Ozone[data$Entity == country] <- mean(neighbor_data$Outdoor_Ozone_Deaths, na.rm = TRUE)
  }
}

spearman_indoor <- cor(data$Indoor_Deaths, data$Neighbor_Indoor, method = "spearman", use = "complete.obs")
spearman_pm <- cor(data$Outdoor_PM_Deaths, data$Neighbor_PM, method = "spearman", use = "complete.obs")
spearman_ozone <- cor(data$Outdoor_Ozone_Deaths, data$Neighbor_Ozone, method = "spearman", use = "complete.obs")

# Print results
cat("Spearman Correlation with Neighbors:\n",
    "Indoor Deaths:", spearman_indoor, "\n",
    "Outdoor PM Deaths:", spearman_pm, "\n",
    "Outdoor Ozone Deaths:", spearman_ozone, "\n")
```
**(This was only done on the subset of countries that had an increasing average trend in deaths per 100000)**
From this data, we get a very low indoor death spearman correation, which suggests that there is a basically no correlation between one country's indoor death rates and its neighbors, which suggests that indoor pollution deaths are heavily influenced by localized factors like household energy sources and individual country policies. (which is to be expected) We get a value of .891 for outdoor deaths, which indicates that there's a high correlation between one country's outdoor deaths and their neighbors, which could be affected by air pollution moving across borders or shared industrial zones. The ozone deaths have a value of 0.5, which indicates some correlation between neighbors and country, which could be attributed to something like similar industrial emissions between countries.


```{r}
for (country in upward_countries) {
  country_data <- data[data$Entity == country, ]
  
  latest_year <- max(country_data$Year, na.rm = TRUE)
  latest_data <- country_data[country_data$Year == latest_year, ]
  
  indoor_deaths <- latest_data$Indoor_Deaths
  neighbor_indoor <- latest_data$Neighbor_Indoor
  
  pm_deaths <- latest_data$Outdoor_PM_Deaths
  neighbor_pm <- latest_data$Neighbor_PM
  
  ozone_deaths <- latest_data$Outdoor_Ozone_Deaths
  neighbor_ozone <- latest_data$Neighbor_Ozone
  
  cat("\nCountry:", country, "\n")
  cat("Year:", latest_year, "\n")
  cat("Indoor Deaths:", indoor_deaths, "Neighbor Average:", neighbor_indoor, "\n")
  cat("Outdoor PM Deaths:", pm_deaths, "Neighbor Average:", neighbor_pm, "\n")
  cat("Outdoor Ozone Deaths:", ozone_deaths, "Neighbor Average:", neighbor_ozone, "\n")
  cat("---------------------------------------------\n")
}

```



```{r}

# Set parameters for permutation
num_iterations <- 1000
permutation_correlations <- matrix(NA, nrow = num_iterations, ncol = 3)
colnames(permutation_correlations) <- c("Indoor", "PM", "Ozone")

# Get list of all countries
all_countries <- unique(data$Entity)

# Calculate observed correlations (from original code)
observed_correlations <- c(Indoor = spearman_indoor, PM = spearman_pm, Ozone = spearman_ozone)
print("Observed Spearman Correlations:")
print(observed_correlations)

set.seed(456) 
for (i in 1:num_iterations) {
  # Shuffle the Entity column
  shuffled_entities <- sample(data$Entity)
  
  shuffled_data <- data
  shuffled_data$Entity <- shuffled_entities
  
  # Recompute Neighbor_Indoor, Neighbor_PM, Neighbor_Ozone with shuffled entities
  shuffled_data$Neighbor_Indoor <- NA
  shuffled_data$Neighbor_PM <- NA
  shuffled_data$Neighbor_Ozone <- NA
  
  for (country in names(neighbors)) {
    neighbor_list <- neighbors[[country]]
    
    # Get shuffled neighbors by matching the shuffled entities
    # Find the positions of the original neighbor names in the shuffled_entities
    # This assumes that neighbor_list contains unique country names
    neighbor_indices <- match(neighbor_list, shuffled_data$Entity)
    valid_indices <- neighbor_indices[!is.na(neighbor_indices)]
    neighbor_shuffled_data <- shuffled_data[valid_indices, ]
    
    if (nrow(neighbor_shuffled_data) > 0) {
      shuffled_data$Neighbor_Indoor[shuffled_data$Entity == country] <- mean(neighbor_shuffled_data$Indoor_Deaths, na.rm = TRUE)
      shuffled_data$Neighbor_PM[shuffled_data$Entity == country] <- mean(neighbor_shuffled_data$Outdoor_PM_Deaths, na.rm = TRUE)
      shuffled_data$Neighbor_Ozone[shuffled_data$Entity == country] <- mean(neighbor_shuffled_data$Outdoor_Ozone_Deaths, na.rm = TRUE)
    }
  }
  
  # Compute Spearman correlations with shuffled neighbors
  perm_indoor <- cor(shuffled_data$Indoor_Deaths, shuffled_data$Neighbor_Indoor, method = "spearman", use = "complete.obs")
  perm_pm <- cor(shuffled_data$Outdoor_PM_Deaths, shuffled_data$Neighbor_PM, method = "spearman", use = "complete.obs")
  perm_ozone <- cor(shuffled_data$Outdoor_Ozone_Deaths, shuffled_data$Neighbor_Ozone, method = "spearman", use = "complete.obs")
  
  permutation_correlations[i, ] <- c(perm_indoor, perm_pm, perm_ozone)
}

p_values <- sapply(1:3, function(j) {
  if (observed_correlations[j] > 0) {
    mean(permutation_correlations[, j] >= observed_correlations[j], na.rm = TRUE)
  } else {
    mean(permutation_correlations[, j] <= observed_correlations[j], na.rm = TRUE)
  }
})

names(p_values) <- c("Indoor", "PM", "Ozone")
print("P-values from Permutation Test:")
print(p_values)

```

```{r}
most_pollution_deaths <- data %>%
  group_by(Entity) %>%
  summarize(
    Total_Deaths_Sum = sum(Total_Deaths, na.rm = TRUE)
  ) %>%
  arrange(desc(Total_Deaths_Sum))

#top 10 countries with the most pollution deaths
head(most_pollution_deaths, 10)

```

9/10 of these countries are considered to be developing countries (all except oceania, which is not really a country but a region)

Now want to look at whether we can see some sort of similarities in the stats between these countries

```{r}
cluster_features <- data[, c("Indoor_Deaths", "Outdoor_PM_Deaths", "Outdoor_Ozone_Deaths")]

cluster_features <- na.omit(cluster_features)

cluster_features_scaled <- scale(cluster_features)

set.seed(123)  # For reproducibility
kmeans_result <- kmeans(cluster_features_scaled, centers = 3)

data$Cluster <- NA
data$Cluster[complete.cases(data[, c("Indoor_Deaths", "Outdoor_PM_Deaths", "Outdoor_Ozone_Deaths")])] <- kmeans_result$cluster

table(data$Entity, data$Cluster)

cluster_summary <- aggregate(
  cluster_features, 
  by = list(Cluster = kmeans_result$cluster), 
  mean
)

cluster_summary

```



```{r}
developing_countries <- c(
  "Afghanistan", "Angola", "Bangladesh", "Benin", "Bhutan", 
  "Bolivia", "Botswana", "Burkina Faso", "Burundi", "Cambodia", 
  "Cameroon", "Central African Republic", "Chad", "Comoros", 
  "Congo", "Cote d'Ivoire", "Democratic Republic of Congo", "Djibouti", 
  "Eritrea", "Ethiopia", "Gambia", "Ghana", "Guatemala", "Guinea", 
  "Guinea-Bissau", "Haiti", "Honduras", "India", "Indonesia", 
  "Kenya", "Kiribati", "Kyrgyzstan", "Laos", "Lesotho", "Liberia", 
  "Madagascar", "Malawi", "Mali", "Mauritania", "Micronesia (country)", 
  "Moldova", "Mozambique", "Myanmar", "Namibia", "Nepal", 
  "Nicaragua", "Niger", "Nigeria", "Pakistan", "Papua New Guinea", 
  "Philippines", "Rwanda", "Samoa", "Sao Tome and Principe", "Senegal", 
  "Sierra Leone", "Solomon Islands", "Somalia", "South Sudan", 
  "Sudan", "Swaziland", "Tajikistan", "Tanzania", "Timor", 
  "Togo", "Uganda", "Uzbekistan", "Vanuatu", "Vietnam", "Yemen", 
  "Zambia", "Zimbabwe"
)

developing_cluster_counts <- table(factor(
  data$Cluster[!is.na(data$Cluster) & data$Entity %in% developing_countries],
  levels = unique(data$Cluster)
))

total_cluster_counts <- table(factor(
  data$Cluster[!is.na(data$Cluster)],
  levels = unique(data$Cluster)
))

table <- rbind(
  Developing = developing_cluster_counts,
  Non_Developing = total_cluster_counts - developing_cluster_counts
)

table
chi_square_test <- chisq.test(table)

chi_square_test
```
Looking at the clustering results, we see that generally developing countries were placed into cluster 1, which is higher amounts of indoor pollution related deaths. Running a chi-squared test on this clustering statistics, we get a p-value of 2.2*10^-6, which is smaller than the significance level 0.05, and thus, we can conclude that the clustering of countries is not independent on whether they are developing or non-developing.

```{r}
data$Cluster <- as.factor(data$Cluster)
anova_deaths <- aov(Total_Deaths ~ Cluster, data = data)
summary(anova_deaths)

tukey_deaths <- TukeyHSD(anova_deaths)
print(tukey_deaths)

boxplot(Total_Deaths ~ Cluster, data = data,
        main = "Comparison of Total Deaths by Cluster",
        xlab = "Cluster",
        ylab = "Total Deaths (per 100k)",
        col = c("blue", "green", "red"))
```

Here we can see how much more deaths in cluster 1 there are than the others. Thus indoor-related pollution seems to be much more deadly.

```{r}
data$Cluster <- as.factor(data$Cluster)
data$Status <- ifelse(data$Entity %in% developing_countries, "Developing", "Developed")

cluster_1_data <- subset(data, Cluster == "1")

boxplot(Total_Deaths ~ Status, data = cluster_1_data,
        main = "Total Deaths in Cluster 1: Developing vs. Developed",
        xlab = "Country Type",
        ylab = "Total Deaths (per 100k)",
        col = c("red", "blue"))
```

However, if we compare the total amount of deaths within cluster 1 by the developed vs developing countries, we see relatively close averages, although the developing countries has significantly more outliers.


```{r}

slopes <- data %>%
  group_by(Entity) %>%
  summarize(
    Cluster = first(Cluster), 
    Slope = coef(lm(Total_Deaths ~ Year))[2]  
  )


anova_slopes <- aov(Slope ~ Cluster, data = slopes)
summary(anova_slopes)

tukey_slopes <- TukeyHSD(anova_slopes)
print(tukey_slopes)


boxplot(Slope ~ Cluster, data = slopes,
        main = "Comparison of Decrease in Deaths Over Time by Cluster",
        xlab = "Cluster",
        ylab = "Slope (Change in Deaths per Year)",
        col = c("blue", "green", "red"))


```


``{r}

risk_factors <- cluster_1_data %>%
  group_by(Status) %>%
  summarize(
    Avg_Indoor_Deaths = mean(Indoor_Deaths, na.rm = TRUE),
    Avg_Outdoor_PM_Deaths = mean(Outdoor_PM_Deaths, na.rm = TRUE),
    Avg_Outdoor_Ozone_Deaths = mean(Outdoor_Ozone_Deaths, na.rm = TRUE)
  )
print(risk_factors)

data$In_Cluster_1 <- ifelse(data$Cluster == "1", 1, 0)


lr_model <- glm(In_Cluster_1 ~ Indoor_Deaths + Outdoor_PM_Deaths + Outdoor_Ozone_Deaths + Status,
                   data = data, family = binomial)
summary(lr_model)
```

Looking at the results from the logistic regression:
Indoor related deaths: for every 1-unit increase in indoor deaths per 100,000, log odds of being in the 1st cluster increases by 0.2123
Outdoor related deaths: For every 1-unit increase in outdoor deaths, odds of being in cluster 1 decrease by 0.146
Ozone: for every 1-unit increase, the log odds of being in cluster 1 decrease by .736
Developing: The odds of being in cluster 1 for a developing country are 2.11 times higher than for developed countries (e^.747..= 2.111)